Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-16 VMCAI-revised-version/final.tex
effectivity

A {\em quasi-ordering} (a qo) is any reflexive and transitive relation $\leq$ over some set $X$ and we often write $(X,\leq)$. 
Given $(X,\leq)$ a quasi-ordering, an {\em upward-closed set} is any set $U \subseteq X$ such that if $x \leq y$ and $x \in U$ then $y \in U $.
A {\em downward-closed set} is any set $D \subseteq X$ such that if $y \leq x$ and $x \in D$ then $y \in D $. 
It is an {\em ideal } if it is also {\em directed}, i.e. it is nonempty and for every $a,b \in D$, there exists $c \in D$ such that $a \leq c$ and $b \leq c$.
To any subset $A \subseteq X$, we may associate
its {\em upward-closure},
 $\uparrow A = \{x \in X \mid \exists a \in A ~ a \leq y\}$
 and its 
 {\em downward-closure},
 $\downarrow A = \{x \in X \mid \exists a \in A ~ y \leq a\}$. 
We abbreviate $\uparrow \{x\}$ (resp. $\downarrow \{x\}$)
as $\uparrow x$ (resp. $\downarrow x$).
%
A {\em basis} of an upward-closed set $I$ is a set $I_b$ such that $I = \uparrow I_b$.

 A {\em well-quasi-ordering} (wqo) is any quasi-ordering $(X,\leq)$ such that, for any infinite sequence $x_0, x_1, x_2, ...$ in $X$, there exist indexes $i \leq j$ with $x_i \leq  x_j$. This property is equivalent to the \emph{finite decomposition} property : every upward-closed set $\emptyset \neq U \subseteq X$ admits a finite basis $B \subseteq X$ such that $U=\uparrow B$.
%
%
%
Wqo admits many other equivalent formulations like : a qo $(X,\leq)$ is a wqo iff $(X,\leq)$ is well founded (i.e. there is no infinite strictly decreasing sequence
of elements of $X$) and $(X,\leq)$ contains no infinite antichains (an antichain is a subset of mutually incomparable elements of $X$). See other equivalences in \cite{schmitz:cel-00727025}. 
%
%
As an example, $(\N^d, \leq)$, the set of vectors of $d$ natural numbers (where $d$ is finite) with component-wise order is a wqo.\\
Quasi-orderings that have no 
infinite antichains
 enjoy a similar \emph{finite decomposition} property than wqo: every downward-closed subset $D \subseteq X$ can be decomposed into a \emph{finite} set of ideals $J_1, J_2, \ldots, J_n$ such that $D = \downarrow (J_1 \cup J_2 \cup \ldots \cup J_n)$ See for example \cite{BFM-ic17}.
%
%
In what follows, a downward-closed set $D$ is represented by its finite set of ideals (or by the minimal elements of its upward-closed complement), and an upward-closed set $U$ is represented by its finite set of minimal elements. \\

@@

Several families of formal models of processes give rise to WSTSs in a natural way with disfferent compatibilities, e.g. compatible counter machines like VASS with $d$ counters and $Q$ a finite set of control-states (and equivalently Petri nets and data Petri nets), reset/transfer VASS are WSTS with strong compatibility for the usual ordering $=\times \leq^d$ on the set of states $S=Q \times \N^d$.
Similarly, lossy channel systems with $d$ channels and $Q$ a finite set of control-states are WSTS (with a non-strong compatibility)
for the ordering $=\times \sqsubseteq^d$ (where $\sqsubseteq$ is the subword ordering on $\Sigma^*$) on $S= Q \times (\Sigma^*)^d$.\\
%%%%

@@

\alain{le rappel ci-dessous est-il entièrement utile ? Je pense qu'on peut l'enlever, et toi ?}
Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.
With the effective pred-basis hypothesis, we obtain:

@@

Let us recall the  definition of vector addition system with (control-)states. 
 \begin{definition} 
A {\em vector addition system with (control-)states (VASS)} in dimension $d$ ($d$-VASS for short) is a finite $\mathds{Z}^d$-labeled directed graph $V = (Q,T)$, where $Q$ is the set of {\em control-states}, and $T \subseteq Q \times \mathds{Z}^d \times Q$ is the set of {\em control-transitions}. 
 \end{definition} 
%
Subsetquently, $Q \times \N^d$ is the set of states of the transition system associated with a $d$-VASS $V$.
For all states $p(\textbf{u}), q(\textbf{v}) \in Q \times \N^d$ and for every control-transition $t = (p, \textbf{z}, q)$, we write $p(\textbf{u}) \xrightarrow{t} q(\textbf{v})$ whenever $\textbf{v} = \textbf{u} + \textbf{z} \geq \textbf{0}$
%
When in the context of a $d$-VASS, we denote $0^d$ by $\textbf{0}$.

A {\em vector addition system (VAS)} in dimension $d$ ($d$-VAS for short) is a $d$-VASS where the set of control-states is a singleton; hence it can be only defined by $T$.\\

@@

In a (not necessarly ordered) transition system $\mathscr{S}=(S,\rightarrow)$, we consider a subset of states $\Safe \subseteq S$, and its complement, $\Bad$.
The \emph{resilience problem} (resp. the \emph{$k$-resilience problem}) for $(\mathscr{S},\Safe)$ is to decide whether from \emph{any} state in 
$S$, \emph{there exists} a path (resp. a path of length smaller than or equal to $k$) that reaches a state in $\Safe$. Resilience is then akin to the Home-Space problem (defined in the introduction) for the set $\Safe$. Resilience can also be viewed as a generalizeation
 of coverability, as
it asks whether for \emph{every} element of $\Bad$ it is possible to cover an element of the basis of $\Safe$.
We use the notation 
$S \longrightarrow^{*} \Safe$ (resp. $S \longrightarrow^{\leq k} \Safe$)
 for 
$\forall x \in S, \exists y \in \Safe$ 
 such that $x \longrightarrow^{*} y$ 
  (resp.  $\forall x \in S, \exists y \in \Safe$ such that $x \longrightarrow^{\leq k} y$).
  In our framework, $\Safe \subseteq S$  is possibly infinite but  must admit a computable finite representation : for example, downward-closed sets and upward-closed sets in wqo and semilinear sets in $\mathbb{N}^d$ have finite representations.

@@

These three resilience problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS rely on the computation of upward or downward-closed sets \cite{DBLP:journals/iandc/AbdullaCJT00,DBLP:journals/tcs/FinkelS01}, we consider upward-closed or downward-closed sets $\Safe$% and $\Bad$
. Since $\Safe \subseteq \pred^*(\Safe)$, one only need to decide whether 
the complement of $\Safe$ is in $\pred^*(\Safe)$. From now on, we use $\Bad$ to denote the complement of $\Safe$.
In \cite{DBLP:journals/corr/abs-2108-00889}, the authors considered that $\Safe$ is upward-closed.  \\

>>>>

A {\em quasi-ordering} (a qo) is any reflexive and transitive relation $\leq$ over some set $X$ and we often write $(X,\leq)$. 
Given a quasi-ordering $(X,\leq)$, an {\em upward-closed set} is any set $U \subseteq X$ such that if $x \leq y$ and $x \in U$ then $y \in U $.
A {\em downward-closed set} is any set $D \subseteq X$ such that if $y \leq x$ and $x \in D$ then $y \in D $. 
It is an {\em ideal } if it is also {\em directed}, i.e. it is nonempty and for every $a,b \in D$, there exists $c \in D$ such that $a \leq c$ and $b \leq c$.
To any subset $A \subseteq X$, we may associate
its {\em upward-closure},
 $\uparrow A = \{x \in X \mid \exists a \in A ~ a \leq x\}$
 and its 
 {\em downward-closure},
 $\downarrow A = \{x \in X \mid \exists a \in A ~ x \leq a\}$. 
We abbreviate $\uparrow \{x\}$ (resp. $\downarrow \{x\}$)
as $\uparrow x$ (resp. $\downarrow x$).
%
A {\em basis} of an upward-closed set $U$ is a set $U_b$ such that $U = \uparrow U_b$; similarly, a {\em basis} of a downward-closed set $D$ is a set $D_b$ such that $D = \downarrow D_b$.

 A {\em well-quasi-ordering} (wqo) is any quasi-ordering $(X,\leq)$ such that, for any infinite sequence $x_0, x_1, x_2, ...$ in $X$, there exist indexes $i \leq j$ with $x_i \leq  x_j$. This property is equivalent to the \emph{finite decomposition} property : every upward-closed set $\emptyset \neq U \subseteq X$ admits a finite basis $B \subseteq X$ such that $U=\uparrow B$.
%
%
%
Wqo admits many other equivalent formulations like : a qo $(X,\leq)$ is a wqo iff $(X,\leq)$ is well founded (i.e. there is no infinite strictly decreasing sequence
of elements of $X$) and $(X,\leq)$ contains no infinite antichains (an antichain is a subset of mutually incomparable elements of $X$). See other equivalences in \cite{DBLP:phd/hal/Halfon18,schmitz:cel-00727025}. 
%
%
As an example, $(\N^d, \leq)$, the set of vectors of $d$ natural numbers (where $d$ is finite) with component-wise order is a wqo.\\
Quasi-orderings that have no 
infinite antichains
 enjoy a similar \emph{finite decomposition} property than wqo: every downward-closed subset $D \subseteq X$ can be decomposed into a \emph{finite} set of ideals $J_1, J_2, \ldots, J_n$ such that $D = J_1 \cup J_2 \cup \ldots \cup J_n$ See for example \cite{BFM-ic17}.
%
%
In what follows, a downward-closed set $D$ is represented by its finite set of ideals (or by the minimal elements of its upward-closed complement), and an upward-closed set $U$ is represented by its finite set of minimal elements. \\

@@

Several families of formal models of processes  \cite{DBLP:journals/tcs/FinkelS01} give rise to WSTSs in a natural way with disfferent compatibilities, e.g. compatible counter machines like VASS with $d$ counters and $Q$ a finite set of control-states (and equivalently Petri nets and data Petri nets), reset/transfer VASS are WSTS with strong compatibility for the usual ordering $=\times \leq^d$ on the set of states $S=Q \times \N^d$.
Similarly, lossy channel systems with $d$ channels and $Q$ a finite set of control-states are WSTS (with a non-strong compatibility)
for the ordering $=\times \sqsubseteq^d$ (where $\sqsubseteq$ is the subword ordering on $\Sigma^*$) on $S= Q \times (\Sigma^*)^d$.\\
%%%%

@@

\alain{le rappel ci-dessous est-il entièrement utile ? Je pense qu'on peut l'enlever, et toi ?}
Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.

Recall the \emph{coverability problem} for ordered transition systems.

\problemx{Coverability problem}
{An ordered transition system $\mathscr{S}=(S,\rightarrow,\longrightarrow)$ and two states $s_0,s \in S$.}
{ $s_0 \in \pred^*(\uparrow s )$ ? \newline}

%An equivalent formulation is : Does there exist $t \in S$ such that $s_0 \longrightarrow^{*} t$ and $s \leq t$ ?
With the effective pred-basis hypothesis, we obtain:

@@

Let us recall the  definition of vector addition system with (control-)states. 
 \begin{definition} 
A {\em vector addition system with (control-)states (VASS)} in dimension $d$ ($d$-VASS for short) is a finite $\mathds{Z}^d$-labeled directed graph $V = (Q,T)$, where $Q$ is the set of {\em control-states}, and $T \subseteq Q \times \mathds{Z}^d \times Q$ is the set of {\em control-transitions}. 
 \end{definition} 
%
Subsetquently, $Q \times \N^d$ is the set of states of the transition system associated with a $d$-VASS $V$.
For all states $p(\textbf{u}), q(\textbf{v}) \in Q \times \N^d$ and for every control-transition $t = (p, \textbf{z}, q)$, we write $p(\textbf{u}) \xrightarrow{t} q(\textbf{v})$ whenever $\textbf{v} = \textbf{u} + \textbf{z} \geq \textbf{0}$.
%
When in the context of a $d$-VASS, we denote $0^d$ by $\textbf{0}$.

A {\em vector addition system (VAS)} in dimension $d$ ($d$-VAS for short) is a $d$-VASS where the set of control-states is a singleton; hence one only needs $T$.\\

@@

In a (not necessarly ordered) transition system $\mathscr{S}=(S,\rightarrow)$, we consider a subset of states $\Safe \subseteq S$, and its complement, $\Bad$.
The \emph{resilience problem} (resp. the \emph{$k$-resilience problem}) for $(\mathscr{S},\Safe)$ is to decide whether from \emph{any} state in 
$S$, \emph{there exists} a path (resp. a path of length smaller than or equal to $k$) that reaches a state in $\Safe$. Resilience is then akin to the Home-Space problem (defined in the introduction) for the set $\Safe$. Resilience can also be viewed as a generalizeation
 of coverability, as
it asks whether for \emph{every} element of $\Bad$ it is possible to cover an element of the basis of $\Safe$.
We use the notation 
$S \longrightarrow^{*} \Safe$ (resp. $S \longrightarrow^{\leq k} \Safe$)
 for 
$\forall x \in S, \exists y \in \Safe$ 
 such that $x \longrightarrow^{*} y$ 
  (resp.  $\forall x \in S, \exists y \in \Safe$ such that $x \longrightarrow^{\leq k} y$).
  In our framework, $\Safe \subseteq S$  is possibly infinite but  must admit a computable finite representation : for example, downward-closed sets and upward-closed sets in wqos and semilinear sets in $\mathbb{N}^d$ have finite representations.

@@

These three resilience problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS rely on the computation of upward or downward-closed sets \cite{DBLP:journals/iandc/AbdullaCJT00,DBLP:journals/tcs/FinkelS01}, we consider upward-closed or downward-closed sets $\Safe$% and $\Bad$
. Since $\Safe \subseteq \pred^*(\Safe)$, one only needs to decide whether 
the complement of $\Safe$ is in $\pred^*(\Safe)$. From now on, we use $\Bad$ to denote the complement of $\Safe$.
In \cite{DBLP:journals/corr/abs-2108-00889}, the authors considered that $\Safe$ is upward-closed.  \\
